---
title: "Final Project Assignment #1"
author: "Kristine Chang and Hanna Kim"
date: "Monday, April 19, 2021"
output: pdf_document
---

```{r global_options, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.height=3, fig.width=5, 
                      fig.align = "center")
library(rmarkdown)
library(tidyverse)
library(mosaic)
library(survminer)
library(tidyr)
library(dplyr)
library(survival)
library(broom)
```

#QUESTIONS
#what to do w missing data in the "anysubstatus" column
#if the variable is already categorical, you could make it have three levels: yes, no, unavailable. if you aren't going to use the info w the third category, you should just avoid the variable.

#group violates PH based on coxzph. what to do ?!?
#looks ok--curves seem parallel
#don't worry about p-value--worry more about understanding the task. find more detail about the test and think about that coxzph test will let you do more variables at a time, is the p-value impacted by using many variables? why would the coxzph give me a sig p-value when the log log curves seem parallel?

#what to write in the "methods" section?
#what i want is YOUR methods. what did you do? can we repeat this? be clear about what the model was, whether you used wald vs likelihood ratio, etc. only tell hardin what you ARE including!

#transformations—how do we know when we should transform a variable? esp a categorical one?
#don't transform a categorical variable
#if you have a numeric variable and you wanna know if the hazard is linear in the age squared (e.g. e^{beta squared}), a one unit increase in age squared would be linear in the log hazard ratio. Age squared would be your unit of measurement
#NOT expecting people to do that

#time dependent covariates...do we need to give a summary of what they are in the paper? is it okay if we only analyze like, one time dependent covariate (maybe education or alcohol or substatus)?
#anderson-gill is on modeling. a way to incorporate a covariate that you think is time-dependent. available in the RMS package
#are we talking about testing for time-dependent covariates, or are we going to use time-dependent covariates? 
#open to an explanation of what it means to be a time-dependent covariate. make it part of the model or dataset--coxzph looks like a wald or something??? if you look carefully at the likelihood we derive in the notes, all of the e^etc need to cancel, and they're not going to cancel if they're a function of t. 

#extension on rough draft,..?

#ask for help for a alcohol vs. anysubstatus graph
#look at hw 8, graph multiple KM curves to compare them to each other
#since we have two categorical variables, graphing multiple KM curves is our best option (esp if we think there is some kind of interaction going on between them)


#TO DO
#should we use anysubstatus? 

#### Load in data set

We load in the data.

```{r}
help <- read_csv("https://pomona.box.com/shared/static/nty3oxpb8db1y763l6uusxtq4gukbcev")
help <- help %>%
  mutate(anysubstatus2 = as.factor(anysubstatus)) %>%
  mutate(alcohol2 = as.factor(alcohol)) %>%
  mutate(group2 = as.factor(group))
```

#### EDA

#variables tried (and not interested): age, gender, race (ask hardin about this one tomorrow morning), M16, E10B1_R, E15C2, G1B_30, abuse2
#variables interested: group (by default), education, anysubstatus, alcohol, F1I 

```{r}
#GROUP

#survival analysis based on group
help_surv <- survfit(Surv(dayslink, linkstatus) ~ group, data=help)
ggsurvplot(help_surv, conf.type = "TRUE") +
  ggtitle("HELP data EDA")

#seeing whether group significantly affects the survival curve
coxph(Surv(dayslink,linkstatus) ~ group, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

#G-statistic
G0 <- 2*(-683.2197-(-718.0145))
1-pchisq(G0,1)

#group does significantly affect the curve

help %>%
  #filter(a9>16) %>%
  group_by(group) %>%
  summarize(n())
```

```{r}
#EDUCATION

#survival analysis based on age
help_surv <- survfit(Surv(dayslink, linkstatus) ~ a9, data=help)
ggsurvplot(help_surv, conf.type = "TRUE") +
  ggtitle("HELP data EDA")

#seeing whether age significantly affects the survival curve
coxph(Surv(dayslink,linkstatus) ~ a9, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

coxph(Surv(dayslink,linkstatus) ~ a9, data = help) %>% tidy()

#G-statistic
G3 <- 2*(-709.6091-(-718.0145))
1-pchisq(G3,1)

#education does significantly affect the curve

help %>%
  filter(a9>5) %>%
  group_by(a9) %>%
  summarize(avgdayslink = mean(dayslink,na.rm=TRUE)) %>%
  ggplot() + 
	geom_line(aes(x = a9, y = avgdayslink), stat="identity") 

#not that many people who have had 16> years education
help %>%
  #filter(a9>16) %>%
  group_by(a9) %>%
  summarize(n())
```

```{r}
#ANYSUBSTATUS

#survival analysis based on substance use
help_surv <- survfit(Surv(dayslink, linkstatus) ~ anysubstatus, data=help)
ggsurvplot(help_surv, conf.type = "TRUE") +
  ggtitle("HELP data EDA")

#seeing whether substance use significantly affects the survival curve
coxph(Surv(dayslink,linkstatus) ~ anysubstatus, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

coxph(Surv(dayslink,linkstatus) ~ as.factor(anysubstatus), data = help) %>% tidy()

#G-statistic
G4 <- 2*(-413.4942-(-718.0145))
1-pchisq(G4,1)

#substance use does significantly affect the curve

#hmm lots of NAs...sus
help %>%
  group_by(anysubstatus) %>%
  summarize(n())
```

```{r}
#F1I: I thought my life had been a failure (0=Rarely/never, 1=Some of the time, 2=Ocassionally, 3=Most of the time)

help_surv <- survfit(Surv(dayslink, linkstatus) ~ f1i, data=help)
ggsurvplot(help_surv, conf.type = "TRUE") +
  ggtitle("HELP data EDA")

coxph(Surv(dayslink,linkstatus) ~ f1i, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

coxph(Surv(dayslink,linkstatus) ~ as.factor(f1i), data = help) %>% tidy()

#G-statistic
G7 <- 2*(-716.0107-(-718.0145))
1-pchisq(G7,1)

#F1I is significant (interesting, I thought those who more often thought their life was a failure would take longer to link with primary care) 
```

```{r}
#alcohol: 1st/2nd drug of choice=Alcohol (0=No, 1=Yes)

help %>%
  group_by(alcohol) %>%
  summarize(n())

help %>%
  group_by(alcohol) %>%
  summarize(avgdayslink = mean(dayslink,na.rm=TRUE)) %>%
  ggplot() + 
	geom_bar(aes(x = alcohol, y = avgdayslink), stat="identity") 

#survival analysis based on alcohol
help_surv <- survfit(Surv(dayslink, linkstatus) ~ alcohol, data=help)
ggsurvplot(help_surv, conf.type = "TRUE") +
  ggtitle("HELP data EDA")

#seeing whether abuse type significantly affects the survival curve
coxph(Surv(dayslink,linkstatus) ~ anysubstatus + alcohol, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ alcohol, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

#G-statistic between anysubstatus + alcohol and just alcohol
G5 <- 2*(--411.8382-(-715.9021))
1-pchisq(G5,1)

#tidy
coxph(Surv(dayslink,linkstatus) ~ alcohol, data = help) %>% tidy()

#alcohol does significantly affect the curve
```

```{r}
#abuse2: Type of abuse (0=No abuse, 1=Physical only, 2=Sexual only, 3=Physical and sexual)

help %>%
  group_by(abuse2) %>%
  summarize(n())

help %>%
  group_by(abuse2) %>%
  summarize(avgdayslink = mean(dayslink,na.rm=TRUE)) %>%
  ggplot() + 
	geom_bar(aes(x = abuse2, y = avgdayslink), stat="identity") 

#survival analysis based on abuse type
help_surv <- survfit(Surv(dayslink, linkstatus) ~ as.factor(abuse2), data=help)
ggsurvplot(help_surv, conf.type="log", censor=F, conf.int=T) +
  ggtitle("HELP data EDA")

#seeing whether abuse type significantly affects the survival curve
coxph(Surv(dayslink,linkstatus) ~ as.factor(abuse2), data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

coxph(Surv(dayslink,linkstatus) ~ as.factor(abuse), data = help) %>% tidy()

#G-statistic
G5 <- 2*(-708.6925-(-718.0145))
1-pchisq(G5,1)

#type of abuse does NOT significantly affect the curve
```

```{r}
coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus) + as.factor(alcohol), data = help) %>% tidy()
coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus) + as.factor(alcohol), data = help) %>% glance()

coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus), data = help) %>% tidy()
coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus), data = help) %>% glance()

#let's keep alcohol
G <- 2*(-386.3138-(-388.8306))
1-pchisq(G,1)

coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + as.factor(anysubstatus) + as.factor(alcohol), data = help) %>% tidy()
coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + as.factor(anysubstatus) + as.factor(alcohol), data = help) %>% glance()

#let's keep a9
G <- 2*(-386.3138-(-389.7752))
1-pchisq(G,1)
```

#r-sections for our paper

#a9: education has a negative coefficient which is unexpected (more years education = lower hazard of being linked to primary care). maybe bc if you're highly educated and in a detox unit, your circumstances are out of the ordinary/extremely strenuous. maybe also shame?

#for alcohol, we see if your 1st/2nd drug of choice is alcohol, you have a higher rate of getting linked to primary care. weird?!?!?!?!? but then you look @ any substatus and see that if you've used any of an assortment of drugs, your rate of getting linked lowers. so, maybe alcohol is relatively less harmful than other substances, so if people's 1st/2nd drug of choice is NOT alcohol, they're using relatively more harmful/illegal/illicit substances. 

#hw 8! plotting KM curves, make 4-6 

#graph 1: survival curve between group and treatment
```{r}
#survival curve between group and treatment
help_surv <- survfit(Surv(dayslink, linkstatus) ~ group, data=help)
ggsurvplot(help_surv, conf.type = "TRUE") +
  ggtitle("HELP data EDA")
```

#group 2: education vs. days to link
```{r}
help %>%
  filter(a9>5) %>%
  group_by(a9) %>%
  summarize(avgdayslink = mean(dayslink,na.rm=TRUE)) %>%
  ggplot() + 
	geom_line(aes(x = a9, y = avgdayslink), stat="identity") +
  labs(title = "Education vs. Days to link to primary care", y = "Average days to link to primary care", x  = "Years of education")
```

#group 3: alcohol + substatus KM curves
```{r}
help_surv2 <- survfit(Surv(dayslink, linkstatus) ~ alcohol2 + anysubstatus2, data=help)

ggsurvplot(help_surv2, conf.type = "TRUE") + 
  ggtitle("KM Curves of alc + substatus")

#maybe do a sexy bar chart ??? alc/sub x mean or median dayslink
help %>%
  group_by(alcohol2, anysubstatus2) %>%
  summarize(avgdayslink = mean(dayslink,na.rm=TRUE)) 

#helppp
`prefers alcohol, hasn't used substances`<-c(201.3793)
`doesn't prefer alcohol, hasn't used substances`<-c(206.8125)
`prefers alcohol, has used substances`<-c(244.8367)
`doesn't prefer alcohol, has used substances`<-c(284.8600)

df <- data.frame(`prefers alcohol, hasn't used substances`,`doesn't prefer alcohol, hasn't used substances`,`prefers alcohol, has used substances`,`doesn't prefer alcohol, has used substances`)

df %>%
  ggplot() + 
	geom_bar(aes(x = a9, y = avgdayslink), stat="identity") +
  labs(title = "Education vs. Days to link to primary care", y = "Average days to link to primary care", x  = "Years of education")

help %>%
ggplot() + 
  geom_bar(aes(x = as.factor(alcohol2), y = dayslink, group = anysubstatus, fill = anysubstatus), 
           position = "dodge", stat = "summary", fun.y = "mean") +
  coord_flip()
```

#final model
```{r}
coxph(Surv(dayslink,linkstatus) ~ as.factor(anysubstatus)*as.factor(alcohol) + as.factor(group) + a9, data = help) %>% tidy()
coxph(Surv(dayslink,linkstatus) ~ as.factor(anysubstatus)*as.factor(alcohol) + as.factor(group) + a9, data = help) %>% glance()

#compared to the additive model, the interactive model above is better
coxph(Surv(dayslink,linkstatus) ~ as.factor(anysubstatus) + as.factor(alcohol) + as.factor(group) + a9, data = help) %>% tidy()
coxph(Surv(dayslink,linkstatus) ~ as.factor(anysubstatus) + as.factor(alcohol) + as.factor(group) + a9, data = help) %>% glance()

G <- 2*(-384.2038-(-386.3138))
1-pchisq(G,1)

#likelihood ratio tests: taking out alcohol
coxph(Surv(dayslink,linkstatus) ~ as.factor(anysubstatus) + as.factor(group) + a9, data = help) %>% tidy()
coxph(Surv(dayslink,linkstatus) ~ as.factor(anysubstatus) + as.factor(group) + a9, data = help) %>% glance()

G <- 2*(-384.2038-(-388.8306))
1-pchisq(G,1)

#likelihood ratio test: taking out education
coxph(Surv(dayslink,linkstatus) ~ as.factor(anysubstatus)*as.factor(alcohol) + as.factor(group), data = help) %>% tidy()
coxph(Surv(dayslink,linkstatus) ~ as.factor(anysubstatus)*as.factor(alcohol) + as.factor(group), data = help) %>% glance()

G <- 2*(-384.2038-(-388.3825))
1-pchisq(G,1)

#likelihood ratio test: taking out substatus
coxph(Surv(dayslink,linkstatus) ~ as.factor(alcohol) + as.factor(group) + a9, data = help) %>% tidy()
coxph(Surv(dayslink,linkstatus) ~ as.factor(alcohol) + as.factor(group) + a9, data = help) %>% glance()

G <- 2*(-384.2038-(-669.8758))
1-pchisq(G,1)

#calculating CIs

#treatment
exp(1.7182155+1.96*0.27459379)
exp(1.7182155-1.96*0.27459379)

#education 
exp(-0.1575370+1.96*0.05499391)
exp(-0.1575370-1.96*0.05499391)

#substance usage
exp(-1.4868904+1.96*0.44711076)
exp(-1.4868904-1.96*0.44711076)

#substance usage x alcohol preference
exp(1.1076608+1.96*0.53036480)
exp(1.1076608-1.96*0.53036480)
```

#checking cox ph assumptions
```{r}
#check PH for interaction b/t alcohol2 and anysubstatus2
help_surv3 <- survfit(Surv(dayslink, linkstatus) ~ alcohol2 + anysubstatus2, data=help)

ggsurvplot(help_surv3, censor=F, conf.int=T, fun="cloglog") +
  ggtitle("HELP data EDA")

#check PH for group2
help_surv4 <- survfit(Surv(dayslink, linkstatus) ~ group2, data=help)

ggsurvplot(help_surv4, censor=F, conf.int=T, fun="cloglog") +
  ggtitle("HELP data EDA")

#make a9 into categorical to check PH...ask Hardin
help <- help %>%
  mutate(a9cat = case_when(
    a9 <= 10 ~ "under10",
    a9 <= 17 ~ "under17",
    TRUE ~ "18-24")) %>%
      mutate(a9cat = factor(a9cat, levels = c("under10","under17","18-24")))

#check PH for a9
help_surv5 <- survfit(Surv(dayslink, linkstatus) ~ a9cat, data=help)

ggsurvplot(help_surv5, censor=F, conf.int=T, fun="cloglog") +
  ggtitle("Log-log graph for a9")

#check all variables
#w/o log transformation of time
cox.zph(coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus) + as.factor(alcohol), data=help))
#w/ log transformation of time
cox.zph(coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus) + as.factor(alcohol), data=help), transform="log")

ggcoxzph(cox.zph(coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus) + as.factor(alcohol), data=help)))
```

