---
title: "Final Project Assignment #1"
author: "Kristine Chang and Hanna Kim"
date: "Monday, April 19, 2021"
output: pdf_document
---

```{r global_options, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, fig.height=3, fig.width=5, 
                      fig.align = "center")
library(rmarkdown)
library(tidyverse)
library(mosaic)
library(survminer)
library(tidyr)
library(dplyr)
library(survival)
library(broom)
```

#QUESTIONS
#what to do w missing data in the "anysubstatus" column
#group violates PH based on coxzph. what to do ?!?
#ask for help for a alcohol vs. anysubstatus graph
#what to write in the "methods" section?
#transformations—how do we know when we should transform a variable? esp a categorical one?
#extension on rough draft,..?
#time dependent covariates...do we need to give a summary of what they are in the paper? is it okay if we only analyze like, one time dependent covariate (maybe education or alcohol or substatus)?

#### Load in data set

We load in the data.

```{r}
help <- read_csv("https://pomona.box.com/shared/static/nty3oxpb8db1y763l6uusxtq4gukbcev")
```

#### EDA

#variables tried (and not interested): age, gender, race (ask hardin about this one tomorrow morning), M16, E10B1_R, E15C2, G1B_30, abuse2
#variables interested: group (by default), education, anysubstatus, alcohol, F1I 

```{r}
#GROUP

#survival analysis based on group
help_surv <- survfit(Surv(dayslink, linkstatus) ~ group, data=help)
ggsurvplot(help_surv, conf.type = "TRUE") +
  ggtitle("HELP data EDA")

#seeing whether group significantly affects the survival curve
coxph(Surv(dayslink,linkstatus) ~ group, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

#G-statistic
G0 <- 2*(-683.2197-(-718.0145))
1-pchisq(G0,1)

#group does significantly affect the curve

help %>%
  #filter(a9>16) %>%
  group_by(group) %>%
  summarize(n())
```

```{r}
#EDUCATION

#survival analysis based on age
help_surv <- survfit(Surv(dayslink, linkstatus) ~ a9, data=help)
ggsurvplot(help_surv, conf.type = "TRUE") +
  ggtitle("HELP data EDA")

#seeing whether age significantly affects the survival curve
coxph(Surv(dayslink,linkstatus) ~ a9, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

coxph(Surv(dayslink,linkstatus) ~ a9, data = help) %>% tidy()

#G-statistic
G3 <- 2*(-709.6091-(-718.0145))
1-pchisq(G3,1)

#education does significantly affect the curve

help %>%
  filter(a9>5) %>%
  group_by(a9) %>%
  summarize(avgdayslink = mean(dayslink,na.rm=TRUE)) %>%
  ggplot() + 
	geom_line(aes(x = a9, y = avgdayslink), stat="identity") 

#not that many people who have had 16> years education
help %>%
  #filter(a9>16) %>%
  group_by(a9) %>%
  summarize(n())
```

```{r}
#ANYSUBSTATUS

#survival analysis based on substance use
help_surv <- survfit(Surv(dayslink, linkstatus) ~ anysubstatus, data=help)
ggsurvplot(help_surv, conf.type = "TRUE") +
  ggtitle("HELP data EDA")

#seeing whether substance use significantly affects the survival curve
coxph(Surv(dayslink,linkstatus) ~ anysubstatus, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

coxph(Surv(dayslink,linkstatus) ~ as.factor(anysubstatus), data = help) %>% tidy()

#G-statistic
G4 <- 2*(-413.4942-(-718.0145))
1-pchisq(G4,1)

#substance use does significantly affect the curve

#hmm lots of NAs...sus
help %>%
  group_by(anysubstatus) %>%
  summarize(n())
```

```{r}
#F1I: I thought my life had been a failure (0=Rarely/never, 1=Some of the time, 2=Ocassionally, 3=Most of the time)

help_surv <- survfit(Surv(dayslink, linkstatus) ~ f1i, data=help)
ggsurvplot(help_surv, conf.type = "TRUE") +
  ggtitle("HELP data EDA")

coxph(Surv(dayslink,linkstatus) ~ f1i, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

coxph(Surv(dayslink,linkstatus) ~ f1i, data = help) %>% tidy()

#G-statistic
G7 <- 2*(-716.0107-(-718.0145))
1-pchisq(G7,1)

#F1I is significant (interesting, I thought those who more often thought their life was a failure would take longer to link with primary care) 
```

```{r}
#alcohol: 1st/2nd drug of choice=Alcohol (0=No, 1=Yes)

help %>%
  group_by(alcohol) %>%
  summarize(n())

help %>%
  group_by(alcohol) %>%
  summarize(avgdayslink = mean(dayslink,na.rm=TRUE)) %>%
  ggplot() + 
	geom_bar(aes(x = alcohol, y = avgdayslink), stat="identity") 

#survival analysis based on alcohol
help_surv <- survfit(Surv(dayslink, linkstatus) ~ alcohol, data=help)
ggsurvplot(help_surv, conf.type = "TRUE") +
  ggtitle("HELP data EDA")

#seeing whether abuse type significantly affects the survival curve
coxph(Surv(dayslink,linkstatus) ~ anysubstatus + alcohol, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ alcohol, data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

#G-statistic between anysubstatus + alcohol and just alcohol
G5 <- 2*(--411.8382-(-715.9021))
1-pchisq(G5,1)

#tidy
coxph(Surv(dayslink,linkstatus) ~ alcohol, data = help) %>% tidy()

#alcohol does significantly affect the curve
```
```{r}
#abuse2: Type of abuse (0=No abuse, 1=Physical only, 2=Sexual only, 3=Physical and sexual)

help %>%
  group_by(abuse2) %>%
  summarize(n())

help %>%
  group_by(abuse2) %>%
  summarize(avgdayslink = mean(dayslink,na.rm=TRUE)) %>%
  ggplot() + 
	geom_bar(aes(x = abuse2, y = avgdayslink), stat="identity") 

#survival analysis based on abuse type
help_surv <- survfit(Surv(dayslink, linkstatus) ~ as.factor(abuse2), data=help)
ggsurvplot(help_surv, conf.type="log", censor=F, conf.int=T) +
  ggtitle("HELP data EDA")

#seeing whether abuse type significantly affects the survival curve
coxph(Surv(dayslink,linkstatus) ~ as.factor(abuse2), data = help) %>% glance()
coxph(Surv(dayslink,linkstatus) ~ 1, data = help) %>% glance()

coxph(Surv(dayslink,linkstatus) ~ as.factor(abuse), data = help) %>% tidy()

#G-statistic
G5 <- 2*(-708.6925-(-718.0145))
1-pchisq(G5,1)

#type of abuse does NOT significantly affect the curve
```

```{r}
coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus) + as.factor(alcohol), data = help) %>% tidy()
coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus) + as.factor(alcohol), data = help) %>% glance()

coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus), data = help) %>% tidy()
coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus), data = help) %>% glance()

#let's keep alcohol
G <- 2*(-386.3138-(-388.8306))
1-pchisq(G,1)

coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + as.factor(anysubstatus) + as.factor(alcohol), data = help) %>% tidy()
coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + as.factor(anysubstatus) + as.factor(alcohol), data = help) %>% glance()

#let's keep a9
G <- 2*(-386.3138-(-389.7752))
1-pchisq(G,1)
```

#r-sections for our paper

#a9: education has a negative coefficient which is unexpected (more years education = lower hazard of being linked to primary care). maybe bc if you're highly educated and in a detox unit, your circumstances are out of the ordinary/extremely strenuous. maybe also shame?

#for alcohol, we see if your 1st/2nd drug of choice is alcohol, you have a higher rate of getting linked to primary care. weird?!?!?!?!? but then you look @ any substatus and see that if you've used any of an assortment of drugs, your rate of getting linked lowers. so, maybe alcohol is relatively less harmful than other substances, so if people's 1st/2nd drug of choice is NOT alcohol, they're using relatively more harmful/illegal/illicit substances. 

#graph 1: survival curve between group and treatment
```{r}
#survival curve between group and treatment
help_surv <- survfit(Surv(dayslink, linkstatus) ~ group, data=help)
ggsurvplot(help_surv, censor=F, conf.int=T, fun="cloglog") +
  ggtitle("HELP data EDA")
```

#group 2: education vs. days to link
```{r}
help %>%
  filter(a9>5) %>%
  group_by(a9) %>%
  summarize(avgdayslink = mean(dayslink,na.rm=TRUE)) %>%
  ggplot() + 
	geom_line(aes(x = a9, y = avgdayslink), stat="identity") +
  labs(title = "Education vs. Days to link to primary care", y = "Average days to link to primary care", x  = "Years of education")
```

#group 3: graph alcohol vs. substatus

#checking cox ph assumptions
```{r}
#checking proportional hazard
cox.zph(coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus) + as.factor(alcohol), data=help))
```

#final model
```{r}
coxph(Surv(dayslink,linkstatus) ~ as.factor(group) + a9 + as.factor(anysubstatus) + as.factor(alcohol), data = help) %>% tidy()
```

